Making Sequential Programs Deterministic

### 2019-04-05 randoop-case-study.tex
More edits in case study section

\subsection{Subject program}

@@

We annotated one package at a time, starting with the most central packges
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the classes and methods had been
documented with respect to their determinism, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we reverse-engineered each
specification, largely from the methods it calls.  Sometimes after we
annotated a subclass, we had to go back and change existing annotation on
the supertype or elsewhere.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.

@@

We reported another suspicious case of order-nondeterminism, in the
\<SpecificationCollection>\-\<.findOverridden> method.  The Randoop developers explained it was
acceptable, after tracing the flow through the program.  (One said, ``This
looks OK to me \ldots\ it took me a little while to decide that.'')  The fact
depended on subtle, undocumented invariants about Randoop that we had not
been able to reverse-engineer.

@@

Java 8 lambdas.

@@

@@

One is that we annotated some core components of Randoop to require all
boolean expressions used in conditionals to be deterministic.  (This is an
option in our implementation, which increases soundness.)  This forced
their clients, and much of the Randoop codebase, to be deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some types could otherwise be @PolyDet, which is the default for
formal parameter and return types.  We plan to re-do the annotation effort
from scratch, and the number of annotations will be much lower.

The second reason is a limitation in \theDeterminismCheckerImplementation.
\TheDeterminismChecker does not support qualifier polymorphism over
classes.  This means that fields must be deterministic.  Our annotations in
Randoop use qualifier polymorphism extensively on methods but cannot use it
on classes, and this leads to many more Det annotations than would be
needed otherwise.

\todo{When we discuss annotation effort, compare to the manual work.
  The annotation effort is
  high, but it's no more than the hundreds of hours that the Randoop
  developers spent, with less success.}

>>>>

\subsection{Subject program}\label{sec:subject-program}

@@

We annotated one package at a time, starting with the most central packges
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change existing annotation on
the supertype or elsewhere because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.

@@

We reported another suspicious case of order-nondeterminism, in the
\<SpecificationCollection>\-\<.findOverridden> method.  The Randoop developers explained it was
acceptable, after tracing the flow through the program.  (One said, ``This
looks OK to me \ldots\ it took me a little while to decide that.'')  The fact
depended on subtle, undocumented invariants about Randoop that we had not
been able to reverse-engineer on our own.

@@

\todo{Java 8 lambdas could help.}

@@

\todo{Easier for someone familiar with the codebase.  Easier on new code.}

It's higher than we would prefer.

@@

One is that we annotated some core components of Randoop to require all
boolean expressions used in conditionals to be deterministic.  (This is an
option in \theDetminismCheckerImplementation, which increases soundness;
see \cref{sec:threats}.)  This forced
their clients --- transitively, much of the Randoop codebase --- to be
annotated as deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some method types could otherwise be qualifier-polymorphic.
This would make them usable in more contexts.  It would also reduce the
number of necessary annotations, since qualifier-polymorphism (the
\<@PolyDet> annotation) is the default for
formal parameter and return types.  We plan to re-do the annotation effort
from scratch, and the number of annotations will be much lower.

The second reason is a limitation in \theDeterminismCheckerImplementation,
which is inherited from the Checker Framework~\cite{checker-framework-tool} upon which
\theDeterminismChecker is built.
The Checker Framework does not support qualifier polymorphism over
classes.  This means that fields must be annotated as deterministic.  Our annotations in
Randoop use qualifier polymorphism extensively on methods but cannot use it
on classes, and this leads to many more \<@Det> annotations than would be
needed otherwise.

Even though the number of annotations is higher for our prototype
implementation than we would prefer, it still compares favorably to the
extensive effort by the Randoop developers (\cref{sec:subject-program}),
especially given that that \theDeterminismChecker found issues that the
developers had not.
